---
title: "MRA Assumptions Exploration"
author: "Efe Aksahin"
date: "2026-01-13"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center"
)
```

# Model building

Linear regression relies on several assumptions. In this project I explore what happens when those assumptions are violated, and more specifically **which estimated quantities change the most** when each violation is introduced in isolation. I focus on:

1. **Linearity**: the conditional mean of the outcome is a linear function of the predictors.
2. **Normality of errors**: the residuals are normally distributed around zero.
3. **Homoscedasticity**: the error variance is constant across fitted values.
4. **Independence of errors**: residuals are uncorrelated across observations.

To study this, I simulate a "perfect" model first and then alter one assumption at a time. For each scenario, I compare the resulting:

- intercept
- regression coefficients for working memory and IQ
- standard errors
- residual standard error
- $R^2$

I also keep one representative dataset from each scenario to visualize how the assumption is violated.

The data generating model is:

$$
\text{math\_test}_i = 22 + 0.2\,\text{IQ}_i + 1.1\,\text{WM}_i + \varepsilon_i,
\qquad i = 1, \dots, n.
$$

with predictors generated as

$$
\text{IQ}_i \stackrel{iid}{\sim} \mathcal{N}(100, 15^2),
\qquad
\text{WM}_i \stackrel{iid}{\sim} \mathcal{N}(26, 8^2),
$$

and under the benchmark model,

$$
\varepsilon_i \stackrel{iid}{\sim} \mathcal{N}(0, 2^2).
$$

So under the ideal case,

$$
\text{math\_test}_i \mid (\text{IQ}_i, \text{WM}_i)
\sim \mathcal{N}\left(22 + 0.2\,\text{IQ}_i + 1.1\,\text{WM}_i, 2^2\right).
$$

```{r packages}
library(ggplot2)
library(dplyr)
library(tidyr)
library(knitr)
library(patchwork)
```

```{r simulation-setup}
set.seed(1334)

n <- 100
num_iterations <- 3000

true_intercept <- 22
true_beta_IQ <- 0.2
true_beta_WM <- 1.1
```

# Perfect model

```{r perfect-simulation}

coefficients_WM <- numeric(num_iterations)
coefficients_IQ <- numeric(num_iterations)
intercepts <- numeric(num_iterations)
standard_errors_WM <- numeric(num_iterations)
standard_errors_IQ <- numeric(num_iterations)
residual_standard_errors <- numeric(num_iterations)
r_squared <- numeric(num_iterations)
p_values_WM <- numeric(num_iterations)
p_values_IQ <- numeric(num_iterations)
for (i in 1:num_iterations) {

  IQ <- rnorm(n,100,15)
  WM <- rnorm(n,26,8)
  error <- rnorm(n,0,2)

  math_test <- true_intercept + true_beta_IQ*IQ + true_beta_WM*WM + error

  data <- data.frame(
    id = 1:n,
    math_test = math_test,
    IQ = IQ,
    WM = WM
  )

  model <- lm(math_test ~ WM + IQ , data = data)

  coefficients_WM[i] <- coef(model)["WM"]
  coefficients_IQ[i] <- coef(model)["IQ"]
  intercepts[i] <- coef(model)[1]
  standard_errors_WM[i] <- summary(model)$coefficients["WM","Std. Error"]
  standard_errors_IQ[i] <- summary(model)$coefficients["IQ","Std. Error"]
  residual_standard_errors[i] <- summary(model)$sigma
  r_squared[i] <- summary(model)$r.squared
  p_values_WM[i] <- summary(model)$coefficients["WM","Pr(>|t|)"]
  p_values_IQ[i] <- summary(model)$coefficients["IQ","Pr(>|t|)"]

  if(i==num_iterations){
    last_data <- data
    last_model <- model
  }
}

parameter_list_perfect <- c(
mean(coefficients_WM),
mean(coefficients_IQ),
mean(intercepts),
mean(standard_errors_WM),
mean(standard_errors_IQ),
mean(residual_standard_errors),
mean(r_squared)
)
```

## Visualizations for the perfect model

```{r fig.width=4, fig.height=3}

p1 <- ggplot(last_data, aes(x = fitted(last_model), y = math_test)) +
geom_point(alpha=.6)+
geom_abline(slope=1,intercept=0,linetype="dashed")+
labs(title="Observed vs Fitted")

p2 <- ggplot(last_data, aes(x=WM,y=math_test))+
geom_point(alpha=.6)+
geom_smooth(method="lm")

p3 <- ggplot(last_data, aes(x=IQ,y=math_test))+
geom_point(alpha=.6)+
geom_smooth(method="lm")

res_df <- data.frame(resid=resid(last_model))

p4 <- ggplot(res_df,aes(x=resid))+
geom_histogram(aes(y=after_stat(density)),bins=20)+
stat_function(fun=dnorm,args=list(mean=mean(res_df$resid),sd=sd(res_df$resid)))

(p1|p2)/(p3|p4)
```

The plots do not show violations of the key assumptions of multiple regression. The dependent variable and the independent variables have a linear relationship, the errors are approximately normal, and have constant variance across the fitted values.

# Violation of Normality

To violate the normality assumption, I generate skewed errors using an exponential distribution.

```{r, fig.width= 4, fig.height=3}

coeff_WM_norm <- numeric(num_iterations)
coeff_IQ_norm <- numeric(num_iterations)
intercepts_norm <- numeric(num_iterations)
se_WM_norm <- numeric(num_iterations)
se_IQ_norm <- numeric(num_iterations)
p_WM_norm <- numeric(num_iterations)
p_IQ_norm <- numeric(num_iterations)
rse_norm <- numeric(num_iterations)
r2_norm <- numeric(num_iterations)

for(i in 1:num_iterations){

  IQ <- rnorm(n,100,15)
  WM <- rnorm(n,26,8)

  error <- rexp(n,1)-1

  math_test <- true_intercept + true_beta_IQ*IQ + true_beta_WM*WM + error

  data_norm <- data.frame(math_test,IQ,WM)

  model_norm <- lm(math_test ~ WM + IQ, data = data_norm)

  coeff_WM_norm[i] <- coef(model_norm)["WM"]
  coeff_IQ_norm[i] <- coef(model_norm)["IQ"]
  intercepts_norm[i] <- coef(model_norm)[1]
  se_WM_norm[i] <- summary(model_norm)$coefficients["WM","Std. Error"]
  se_IQ_norm[i] <- summary(model_norm)$coefficients["IQ","Std. Error"]
  p_WM_norm[i] <- summary(model_norm)$coefficients["WM","Pr(>|t|)"]
  p_IQ_norm[i] <- summary(model_norm)$coefficients["IQ","Pr(>|t|)"]
  rse_norm[i] <- summary(model_norm)$sigma
  r2_norm[i] <- summary(model_norm)$r.squared

  if(i==num_iterations){
    last_data_norm <- data_norm
    last_model_norm <- model_norm
  }
}
```

```{r, fig.width= 4, fig.height=3}

res_norm <- data.frame(resid=resid(last_model_norm))

ggplot(res_norm,aes(x=resid))+
geom_histogram(aes(y=after_stat(density)),bins=20)+
labs(title="Residual distribution under non-normal errors")

```

Even though the residual distribution is clearly skewed, the regression coefficients remain close to the true values. This illustrates that **normality mainly affects inference rather than the coefficient estimates themselves**.

# Violation of Homoscedasticity

Now the variance of the errors increases with working memory.

```{r, fig.width= 4, fig.height=3}

coeff_WM_het <- numeric(num_iterations)
coeff_IQ_het <- numeric(num_iterations)
intercepts_het <- numeric(num_iterations)
se_WM_het <- numeric(num_iterations)
se_IQ_het <- numeric(num_iterations)
p_WM_het <- numeric(num_iterations)
p_IQ_het <- numeric(num_iterations)
rse_het <- numeric(num_iterations)
r2_het <- numeric(num_iterations)

for(i in 1:num_iterations){

  IQ <- rnorm(n,100,15)
  WM <- rnorm(n,26,8)

  error_sd <- 0.4 + 0.08*WM
  error <- rnorm(n,0,error_sd)

  math_test <- true_intercept + true_beta_IQ*IQ + true_beta_WM*WM + error

  data_het <- data.frame(math_test,IQ,WM)

  model_het <- lm(math_test ~ WM + IQ, data = data_het)

  coeff_WM_het[i] <- coef(model_het)["WM"]
  coeff_IQ_het[i] <- coef(model_het)["IQ"]
  intercepts_het[i] <- coef(model_het)[1]
  se_WM_het[i] <- summary(model_het)$coefficients["WM","Std. Error"]
  se_IQ_het[i] <- summary(model_het)$coefficients["IQ","Std. Error"]
  p_WM_het[i] <- summary(model_het)$coefficients["WM","Pr(>|t|)"]
  p_IQ_het[i] <- summary(model_het)$coefficients["IQ","Pr(>|t|)"]
  rse_het[i] <- summary(model_het)$sigma
  r2_het[i] <- summary(model_het)$r.squared

  if(i==num_iterations){
    last_data_het <- data_het
    last_model_het <- model_het
  }
}

```

```{r, fig.width= 4, fig.height=3}

ggplot(last_data_het,aes(x=fitted(last_model_het),y=resid(last_model_het)))+
geom_point(alpha=.6)+
geom_hline(yintercept=0,linetype="dashed")+
labs(title="Residual funnel pattern")

```

The funnel shape indicates increasing variance. Again, the coefficients remain largely unbiased, but **standard errors are affected because the model assumes constant variance**.

# Violation of Independence

Autocorrelation is introduced by making each residual depend on the previous residual.

```{r}

coeff_WM_auto <- numeric(num_iterations)
coeff_IQ_auto <- numeric(num_iterations)
intercepts_auto <- numeric(num_iterations)
se_WM_auto <- numeric(num_iterations)
se_IQ_auto <- numeric(num_iterations)
p_WM_auto <- numeric(num_iterations)
p_IQ_auto <- numeric(num_iterations)
rse_auto <- numeric(num_iterations)
r2_auto <- numeric(num_iterations)

for(i in 1:num_iterations){

  IQ <- rnorm(n,100,15)
  WM <- rnorm(n,26,8)

  u <- rnorm(n,0,2)
  error <- numeric(n)

  error[1] <- u[1]

  for(j in 2:n){
    error[j] <- 0.65*error[j-1] + u[j]
  }

  math_test <- true_intercept + true_beta_IQ*IQ + true_beta_WM*WM + error

  data_auto <- data.frame(math_test,IQ,WM)

  model_auto <- lm(math_test ~ WM + IQ, data = data_auto)

  coeff_WM_auto[i] <- coef(model_auto)["WM"]
  coeff_IQ_auto[i] <- coef(model_auto)["IQ"]
  intercepts_auto[i] <- coef(model_auto)[1]
  se_WM_auto[i] <- summary(model_auto)$coefficients["WM","Std. Error"]
  se_IQ_auto[i] <- summary(model_auto)$coefficients["IQ","Std. Error"]
  p_WM_auto[i] <- summary(model_auto)$coefficients["WM","Pr(>|t|)"]
  p_IQ_auto[i] <- summary(model_auto)$coefficients["IQ","Pr(>|t|)"]
  rse_auto[i] <- summary(model_auto)$sigma
  r2_auto[i] <- summary(model_auto)$r.squared

  if(i==num_iterations){
    last_model_auto <- model_auto
  }
}
```

```{r, fig.width= 4, fig.height=3}

res_auto <- data.frame(
order=1:n,
res=resid(last_model_auto)
)

ggplot(res_auto,aes(x=order,y=res))+
geom_line()+
geom_hline(yintercept=0,linetype="dashed")+
labs(title="Autocorrelated residuals")

```

Residuals now show a clear pattern over observations, violating independence.

# Violation of Linearity

Finally, I introduce a quadratic relationship with working memory.

```{r, fig.width= 4, fig.height=3}

coeff_WM_nonlin <- numeric(num_iterations)
coeff_IQ_nonlin <- numeric(num_iterations)
intercepts_nonlin <- numeric(num_iterations)
se_WM_nonlin <- numeric(num_iterations)
se_IQ_nonlin <- numeric(num_iterations)
p_WM_nonlin <- numeric(num_iterations)
p_IQ_nonlin <- numeric(num_iterations)
rse_nonlin <- numeric(num_iterations)
r2_nonlin <- numeric(num_iterations)

for(i in 1:num_iterations){

  IQ <- rnorm(n,100,15)
  WM <- rnorm(n,26,8)

  error <- rnorm(n,0,2)

  math_test <- true_intercept +
    true_beta_IQ*IQ +
    true_beta_WM*WM -
    0.03*(WM-26)^2 +
    error

  data_nonlin <- data.frame(math_test,IQ,WM)

  model_nonlin <- lm(math_test ~ WM + IQ, data = data_nonlin)

  coeff_WM_nonlin[i] <- coef(model_nonlin)["WM"]
  coeff_IQ_nonlin[i] <- coef(model_nonlin)["IQ"]
  intercepts_nonlin[i] <- coef(model_nonlin)[1]
  se_WM_nonlin[i] <- summary(model_nonlin)$coefficients["WM","Std. Error"]
  se_IQ_nonlin[i] <- summary(model_nonlin)$coefficients["IQ","Std. Error"]
  p_WM_nonlin[i] <- summary(model_nonlin)$coefficients["WM","Pr(>|t|)"]
  p_IQ_nonlin[i] <- summary(model_nonlin)$coefficients["IQ","Pr(>|t|)"]
  rse_nonlin[i] <- summary(model_nonlin)$sigma
  r2_nonlin[i] <- summary(model_nonlin)$r.squared

  if(i==num_iterations){
    last_data_nonlin <- data_nonlin
  }
}

```

```{r, fig.width= 4, fig.height=3}

ggplot(last_data_nonlin,aes(x=WM,y=math_test))+
geom_point(alpha=.6)+
geom_smooth()+
geom_smooth(method="lm",linetype="dashed")+
labs(title="True nonlinear relationship")

```

Here the linear regression tries to approximate a curved relationship with a straight line. This leads to **biased slope estimates**, making this assumption violation particularly problematic.

# Overall Conclusion

```{r}
comparison_table <- data.frame(
  Scenario = c("Perfect Model",
               "Non-normal Errors",
               "Heteroscedastic Errors",
               "Autocorrelated Errors",
               "Nonlinear Relationship"),

  Intercept = c(
    mean(intercepts),
    mean(intercepts_norm),
    mean(intercepts_het),
    mean(intercepts_auto),
    mean(intercepts_nonlin)
  ),

  WM_Coefficient = c(
    mean(coefficients_WM),
    mean(coeff_WM_norm),
    mean(coeff_WM_het),
    mean(coeff_WM_auto),
    mean(coeff_WM_nonlin)
  ),

  IQ_Coefficient = c(
    mean(coefficients_IQ),
    mean(coeff_IQ_norm),
    mean(coeff_IQ_het),
    mean(coeff_IQ_auto),
    mean(coeff_IQ_nonlin)
  ),

  WM_SE = c(
    mean(standard_errors_WM),
    mean(se_WM_norm),
    mean(se_WM_het),
    mean(se_WM_auto),
    mean(se_WM_nonlin)
  ),

  IQ_SE = c(
    mean(standard_errors_IQ),
    mean(se_IQ_norm),
    mean(se_IQ_het),
    mean(se_IQ_auto),
    mean(se_IQ_nonlin)
  ),

  WM_pvalue = c(
    mean(p_values_WM),
    mean(p_WM_norm),
    mean(p_WM_het),
    mean(p_WM_auto),
    mean(p_WM_nonlin)
  ),

  IQ_pvalue = c(
    mean(p_values_IQ),
    mean(p_IQ_norm),
    mean(p_IQ_het),
    mean(p_IQ_auto),
    mean(p_IQ_nonlin)
  ),

  Residual_SE = c(
    mean(residual_standard_errors),
    mean(rse_norm),
    mean(rse_het),
    mean(rse_auto),
    mean(rse_nonlin)
  ),

  R_squared = c(
    mean(r_squared),
    mean(r2_norm),
    mean(r2_het),
    mean(r2_auto),
    mean(r2_nonlin)
  )
)

comparison_table[-1] <- round(comparison_table[-1], 3)

kable(
  comparison_table,
  caption = "Average estimated parameters across 3,000 simulations per scenario (true values: intercept = 22, b_WM = 1.1, b_IQ = 0.2)",
  col.names = c("Scenario", "Intercept", "Coef (WM)", "Coef (IQ)",
                "SE (WM)", "SE (IQ)", "p (WM)", "p (IQ)",
                "Residual SE", "R-sq"),
  align = c("l", rep("r", 9))
)
```

The simulation illustrates that not all regression assumptions affect the model in the same way.

Violations of **normality**, **homoscedasticity**, and **independence** mainly affect the **uncertainty estimates** of the model. The regression coefficients themselves remain close to their true values across all three scenarios, indicating that OLS estimation is relatively robust to these violations in terms of bias. However, the standard errors become unreliable: heteroscedasticity inflates them for the affected predictor (WM), while autocorrelation leads to underestimated standard errors, making the model appear more precise than it actually is. Non-normal errors have the least visible impact on the standard errors, consistent with the central limit theorem providing some protection at $n = 100$.

In contrast, violations of **linearity** directly distort the estimated regression coefficients. Because the model forces a straight line through a curved relationship, the WM slope is pulled away from the true value — this is not a matter of noisy estimation but systematic bias. No amount of additional data would fix this; the model is simply misspecified.

From an applied perspective, this means the consequences of each violation differ in kind, not just degree. Normality, homoscedasticity, and independence violations call for corrections to inference (e.g. robust standard errors, GLS), while a linearity violation requires reconsidering the model structure itself. **Among all the assumptions explored here, misspecification of the mean function is the most damaging because it compromises the coefficients that are typically the primary target of interpretation.**
